---
title: "Report on Motorcycle Accident Deaths in Colombia"
subtitle: "Tidyverse and RMarkdown Practice"
author: "YOUR NAME HERE" # Remember to add your name!
output:
  prettydoc::html_pretty:
    theme: architect
date: "`r Sys.Date()`"
---

```{r include = FALSE}
# Set chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
# TASK 1: LOAD THE REQUIRED PACKAGES
if(!require(pacman)) install.packages("pacman")
"WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 2: IMPORT THE DATA
col_accidents <- "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 3: CLEAN THE COLUMN NAMES 
col_accidents_clean <- col_accidents %>% 
  "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 4: RENAME AND TRANSFORM VARIABLES
col_accidents_clean2 <- col_accidents_clean %>%
  "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 5: CREATE A SUBSET
col_accidents_subset <- "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 6: SUMMARIZE DATA
age_sex_count <- "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 7: VISUALIZE SUMMARY DATA
"WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 8: EXPORT YOUR OUTPUTS
"WRITE_YOUR_CODE_HERE"
```

# Introduction

The dataset analysed in this report provides information about **`r nrow(col_accidents_subset)`** deaths caused by motorcycle accidents in Medellín, a Colombian city, from 2012 to 2015. The data is sourced from medical and police certificates and can be accessed [here](https://zenodo.org/record/4836304).

# Age-sex distribution of motorcyle deaths

In this analysis, we looked at the number of deaths per age and sex group.

```{r}
# Print your graph here.

```

```{r}
# TASK 9: ADD INLINE CODE AND COMMENTARY 

## Each of the 3 statistics in this code chunk will be quoted in the text below.

max_sex_group <- age_sex_count %>% 
  filter(n == max(n)) %>% 
  pull(gender)

max_age_group <- age_sex_count %>% 
  "WRITE_YOUR_CODE_HERE"

max_n_deaths <- "WRITE_YOUR_CODE_HERE"
```

The age group with the most fatal accidents was `r paste0(max_sex_group, "s")` aged `"WRITE_YOUR_INLINE_CODE_HERE"`, with `"WRITE_YOUR_INLINE_CODE_HERE"` deaths.

## Analysis of Driver Position 

We further analyzed the position of the driver involved in these accidents for the age-sex group with the most fatal accidents. 

The table below shows the breakdown of accidents by driver position:

```{r}
# TASK 10: SUMMARIZE DIRVER POSITION DATA 
"WRITE_YOUR_CODE_HERE"
```
